% -*- LaTeX -*-

%\begin{abstract}
 %  After sound, images and videos, 3D models are everywhere and used for multimedia, video games, numerical simulations, manufacturing, computer-aided medicine, culturage heritage and other fields. In parallel to this new multimedia revolution during the past decade, exceptional progress was made in elaborating the theoretical and algorithmic foundations of 3D geometric modeling, and providing efficient algorithms and codes for applications such as surface reconstruction, mesh generation and point cloud processing.  Extending those techniques to higher dimensions is a grand challenge with a huge potential impact in science and engineering but is currently extremely limited and asks for new algorithmic  breakthrough. This project aims at settling the algorithmic foundations of geometric modeling in higher dimensions and to propose a well-principled ground-breaking software platform allowing technological advances for varied applications in science and engineering.


\paragraph{Proposal summary.} 

% <<<<<<< .mine
% =======
% The need for computer-aided {\em understanding} of geometric structures, geometry understanding for short, is ubiquitous in science and has become an essential part of {\em scientific computing} and {\em data analysis}. Geometry understanding is by no means limited to three dimensions and many applications in physics, biology, and engineering require a keen understanding of the geometry of a variety of higher dimensional spaces. Let us mention phase space in particle physics, invariant manifolds in dynamical systems, configuration spaces of mechanical systems, conformational spaces of molecules, image manifolds, and shape spaces, to name a few. In {\em data analysis and manifold learning}, data are often thought as points in some high-dimensional metric space. Those points are usually not uniformly distributed in the embedding space but lie close to some {\em low-dimensional manifold}, which reflects the fact that the physical system that produced the data has a moderate number of degrees of freedom.  Understanding the geometry of this manifold is key to understanding the underlying system. We do not refer to a single action: the term geometry understanding encompasses a collection of tasks including representing and approximating the manifold, extracting the manifold from the point set, and inferring some geometric or topological properties.
% >>>>>>> .r121

The central goal of this proposal is to settle the algorithmic
foundations of geometry understanding in dimensions higher than 3.  We
coin the term {\em geometry understanding} to encompass a collection
of tasks including the computer representation and the approximation
of geometric structures, and the inference of geometric or topological
properties from sampled
shapes.  % geometric structure from point sets.

The need to understand geometric structures is ubiquitous in science
and has become an essential part of {scientific computing} and
{data analysis}. In {\em scientific computing},
geometry understanding is by no means limited to
three dimensions and many applications in physics, biology, and
engineering require a keen understanding of the geometry of a variety
of {\em higher dimensional spaces}. Let us mention phase space in particle
physics, invariant manifolds in dynamical systems, configuration
spaces of mechanical systems, conformational spaces of molecules,
image manifolds, and shape spaces, to name a few.  {\em Data
  analysis} is another place where understanding
geometry in high dimensional spaces is  needed to
 capture concise information from the underlying structure of
 data. Our approach is complementary to {\em manifold learning}
 techniques and aims at  developing an effective theory for {\em geometric and
 topological data analysis.}
% The central focus of this proposal is the computer analysis of
% geometric structures, which we refer to as {\em geometry
%   understanding}.  The need for analyzing geometric structures is
% ubiquitous in science and has become an essential part of {\em scientific
% computing and data analysis}. Geometry understanding is by no means
% limited to 3-dimensions and many applications in physics, biology and
% engineering require a keen understanding of the geometry of a variety
% of higher dimensional spaces. Let us mention phase space in particle
% physics, invariant manifolds in dynamical systems, configuration
% spaces of mechanical systems, conformational spaces of molecules,
% image manifolds, or shape spaces.
% Data analysis  is another place where understanding
% geometry in high-dimensional spaces is needed to
% capture concise information from the underlying structure of the data.  
% Since data are produced at an unprecedented rate in all
% sciences, understanding the geometry of high-dimensional data %point clouds
% has become a core task in science and engineering.

The central goal of this proposal is to settle the {\em algorithmic
  foundations} of geometry understanding in {\em dimensions higher
  than 3}.  We intend to develop {\em scalable representations} in the
form of {\em simplicial complexes}, and {\em practical algorithms} to
approximate highly nonlinear shapes, and to infer geometric and
topological properties from data subject to significant {\em defects}
and under {\em realistic conditions}.  As is common in many
applications across science and engineering, we will assume that the
objects of interest can be modeled as {\em low-dimensional manifolds}
embedded in possibly high-dimensional spaces. By exploiting the {\em
  intrinsic properties} of the objects, we will produce intrinsic
dimension-sensitive data structures and algorithms that will break the
current computational bottleneck.

To reach these objectives, the guiding principle will be to foster a
symbiotic relationship between theory and practice, and to address
{\em fundamental research} issues along three parallel advancing
fronts. We will simultaneously develop {\em mathematical approaches}
providing theoretical guarantees, {\em effective algorithms} that are
amenable to theoretical analysis and rigorous experimental validation,
and {\em perennial software} development.  We will undertake the
development of a high-quality open source {\em software platform} to
implement the most important geometric data structures and algorithms
at the heart of geometry understanding in higher dimensions. The
platform will be a unique vehicle towards researchers from other
fields and will serve as a {\em basis for groundbreaking advances} in
scientific computing and data analysis.


%\vspace{-3mm}

\paragraph{Keywords :} Computational geometry, computational topology,
manifold learning, homology inference, simplicial complexes,
scientific computing, data analysis
