\section{Extended synopsis of the project}

\paragraph{Geometry understanding in higher-dimensions.} 
% The central focus of this proposal is the %extraction, representation and
% computer analysis of geometric structures, which we refer to as {\em geometry understanding}.  
% The need for {\em approximating} complex shapes is ubiquitous in science and has become an essential part of {\em scientific computing}. It is by no means limited to 3-dimensions and
% many applications in physics, biology and engineering require a keen understanding of the geometry of a variety of higher dimensional spaces. Let us mention phase space in particle physics, invariant manifolds in dynamical systems, configuration spaces of mechanical systems, conformational spaces of molecules, image manifolds, shape spaces, to name a few.  % A problem of central importance in data analysis is to recover meaningful low-dimensional structures hidden in high-dimensional data.
% In {\em data analysis}, data are often thought as points in some  high-dimensional metric space~\cite{dld-hdda-2000}. Those points are usually not uniformly distributed in the embedding space but 
% %due to the very nature of the system that produced those data, 
% lie close to some {\em low-dimensional manifold}, which reflects the fact that the physical system that produced the data has a moderate number of  degrees of freedom. {\em Extracting}  an approximation of  the manifold  from the data,   {\em inferring some of its geometric and topological properties} are key to understanding the underlying system. % Extracting, representing, approximating, analyzing shapes and infering some of their geometric and topological features 
% These are fundamental questions in % under the umbrella of 
% {\em geometry understanding.}


% <<<<<<< .mine
% The need for computer-aided {\em understanding} of geometric structures, geometry understanding for short, is ubiquitous in science and has become an essential part of {\em scientific computing} and {\em data analysis}. Geometry understanding is by no means limited to 3-dimensions and many applications in physics, biology and engineering require a keen understanding of the geometry of a variety of higher dimensional spaces. Let us mention phase space in particle physics, invariant manifolds in dynamical systems, configuration spaces of mechanical systems, conformational spaces of molecules, image manifolds, shape spaces, to name a few. In {\em data analysis and manifold learning}, data are often thought as points in some high-dimensional metric space~\cite{dld-hdda-2000}. Those points are usually not uniformly distributed in the embedding space but lie close to some {\em low-dimensional manifold}, which reflects the fact that the physical system that produced the data has a moderate number of degrees of freedom.  Understanding the geometry of this manifold is key to understanding the underlying system. We do not refer to a single action and geometry understanding encompases a collection of problems like representing and approximating the manifold, extracting the manifold from the point set, inferring some geometric or topological properties.
% =======
The need for computer-aided {\em understanding} of geometric structures, geometry understanding for short, is ubiquitous in science and has become an essential part of {\em scientific computing} and {\em data analysis}. Geometry understanding is by no means limited to three dimensions and many applications in physics, biology, and engineering require a keen understanding of the geometry of a variety of higher dimensional spaces. Let us mention phase space in particle physics, invariant manifolds in dynamical systems, configuration spaces of mechanical systems, conformational spaces of molecules, image manifolds, and shape spaces, to name a few. In {\em data analysis and manifold learning}, data are often thought as points in some high-dimensional metric space~\cite{dld-hdda-2000}. Those points are usually not uniformly distributed in the embedding space but lie close to some {\em low-dimensional manifold}, which reflects the fact that the physical system that produced the data has a moderate number of degrees of freedom.  Understanding the geometry of this manifold is key to understanding the underlying system. We do not refer to a single action: the term geometry understanding encompases a collection of tasks including representing and approximating the manifold, extracting the manifold from the point set, and inferring some geometric or topological properties.


Let us illustrate our motivation and objectives through the paradigmatic example of {\em energy landscapes of molecules}. Understanding  energy landscapes is a major challenge in chemistry and biology but,  despite a lot of efforts and a wide variety of approaches, little is understood about the actual structures underlying such landscapes. The case of cyclo-octane $C_8H_{16}$, a cyclic alkane used in manufacture of plastics, is instructive. This relatively simple molecule has been studied in chemistry for over 40 years but it is only very recently that its conformational space has been fully understood. By analyzing a dataset of 1M points in $\R^{72}$, each describing a cyclo-octane conformation, it has been shown that the conformation space of cyclo-octane has the unexpected geometry of  a multi-sheeted 2-dimensional surface composed of a sphere and a Klein bottle, intersecting in two rings~\cite{mtcw-tco-2010}.  Besides its fundamental interest, such a discovery opens new avenues for understanding the energy landscape of cyclo-octane. Extending this type of analysis to large molecules, in particular to proteins, would have tremendous implications. Many other examples with high potential remain to be solved in domains as varied as neurosciences, medical imaging, speech recognition and astrophysics. {\em The crux is to have access to robust and efficient geometric data structures and algorithms to understand geometry in higher dimensions}. This is the grand challenge the Gudhi project wants to take up.



\paragraph{The curses of higher-dimensional geometry.} 
Many difficulties have to be faced when processing and analyzing
high-dimensional geometries. First, the dimensionality severely restricts our intuition and ability to visualize the data.  Understanding higher dimensional shapes must hence rely on {\em automated} methods and tools that produce {\em provably correct} results under {\em realistic circumstances}.

Second, major difficulties come from the fact that the complexity of data structures and algorithms used to approximate shapes rapidly grows as the dimensionality increases, which makes them intractable in high dimensions.  This phenomenon, referred to as the {\em curse of dimensionality}, prevents  in particular from subdividing the ambient space, as is usually done in 3-space, since the size of any such subdivision depends exponentially on the ambient dimension. Instead, any practical method must be {\em sensitive to the intrinsic dimension} (usually unknown) of the shape under analysis. As already mentionned, and observed in the cyclo-octane example, the intrinsic dimension can often be assumed to be much lower than the ambient dimension. This is a powerful assumption we refer to as the {\em manifold model}.\footnote{We use the term manifold in a rather liberal way, including stratified manifolds.} %  Note that, in the  cyclo-octane example above, the object of interest is 2-dimensional and lives in a much higher-dimensional ambient space.
% but can be modeled as a {\em low-dimensional manifold}.% This is a common assumption  in many applications  that captures the fact that, most of the time, high-dimensional data are associated to physical systems that have a moderate number of  degrees of freedom.%parameters.


In addition, high-dimensional data often suffer from significant {\em defects}, including sparsity, noise, and outliers that may hide the intrinsic dimension of the underlying structure. This is particularly so in the case of biological data, such as high throughput data from microarray or other sources. % Moreover, the structure and occurrence of geometric features in the data may depend on the {\em scale} at which it is considered, thus requiring the analysis to automatically select the appropriate scale.  These issues have been considered in the statistical community but not to the same extent in a geometric framework.

\paragraph{The emergence of geometry understanding in higher dimensions.}
The last decade has seen tremendous progress in  the understanding of geometry in high-dimensional spaces. %  In robotics~\cite{sml-pa-2006}, randomized techniques have been proposed to capture the topology of configuration spaces and to search paths.
In signal and image processing, and in machine learning, a variety of techniques, known as {\em nonlinear dimensionality reduction} have been proposed to reduce the dimension of data, to learn nonlinear manifolds and to cluster data~\cite{lv-nldr-2007}. % In computational geometry, new approaches have been proposed to solve basic problems like searching nearest neighbours, computing smallest enclosing ellipsoids or approximating convex sets~\cite{hp-gaa-2011}.
Such techniques are widely used but have limited guarantees and 
%address elementary questions (which may be very hard to solve though) and 
impose strong constraints on the dimension or topology of the shapes they can successfully handle. 

The techniques developed in {\em computational geometry and topology}  are complementary. They aim at processing and analyzing shapes with non trivial geometry and topology~\cite{hh-ct-2010}. 
Emblematic problems such as mesh generation and surface reconstruction in 3-dimensions are now well-understood and several provably correct and highly efficient solutions are now available~\cite{geometrica-ecg-book}. 
%The concepts of $\varepsilon$-samples, restricted Delaunay triangulation, anisotropic meshes %emerged together with the first efficient and provably correct algorithms for 

Attempts to analyze higher dimensional shapes led to the development of beautiful pieces of theory with deep roots in various areas of mathematics like Riemannian geometry, geometric measure theory, differential and algebraic topology. Let us mention  the emergence of a sampling theory of geometric objects, and of geometric inference~\cite{geometrica-ccl09}, and the groundbreaking invention and rapid growth of persistent homology~\cite{eh-ph-2008}.
These advances 
attracted  interest in several fields like biological data analysis~\cite{fpgo-airc-2009}, computer vision~\cite{cids-lbsni-2008} and sensor networks~\cite{rg-bptd-2008}. However, 
until now, the applications have been limited  to rather simple cases and to low dimensions.


\paragraph{The grand challenge~: settling the algorithmic foundations.}
Since many of the most promising applications are in higher dimensions, there is a pressing need to elaborate more effective tools that scale to real problems.  We identify the lack of  algorithmic foundations for geometry understanding in higher dimensions as 
the main cause of the current limited impact of geometric and topological methods.  Settling such foundations
%al geometric modeling 
is a challenge of great theoretical and practical significance at the heart of the Gudhi~project.


A {\em tenet} of this proposal is that, to take up the challenge, we need a {\em global approach} involving tight and long-standing interactions between {\em mathematical research, algorithmic design} and {\em advanced software development}. We believe that this is key to obtaining methods with built-in robustness, scalability and guarantees, which are the best promises for {\em impact in the long run.}
% Such a global approach has been successfully carried out in low dimensions with the 
% development of the Computational Geometric Algorithms Library CGAL~\cite{cgal}. 

%We want to build upon this success 
We strongly believe that, by following this paradigm,  our ambitious objective is realistic and can be reached. To pave the way towards this goal, we have identified  four main scientific challenges.

% \paragraph{The emerging field of structural data analysis.} {\em Early geometric approaches, 
% dimension reduction, NL manifold learning, computational topology, geometric inference
% 3d data processing, surface reconstruction, simplification. }


\paragraph{Scientific challenge 1 :  Choosing the right representation.}
%Going beyond affine models and Euclidean spaces.}

% Standard methods like PCA in Machine Learning to deal with high dimensional data assume that the data can be well approximated by some affine subspace of small dimension.  To overcome this 

As discussed above, dimensionality reduction techniques cannot provide precise approximation of complicated shapes (as required in scientific computing) nor compute essential features of a shape like its topological invariants.
% In the last decades, a set of new geometric methods, known as manifold learning, have been developed with the intent of parametrizing nonlinear shapes embedded in high-dimensional spaces. Although widely used, those methods  assume very restrictive hypotheses on the geometry of the manifolds sampled by the datapoints to ensure correctness. 
More expressive representations of shapes are provided % , inspired by what has been done in 3-dimensions, consists in approximating
by {\em simplicial complexes}, the analogue of triangulations in higher dimensions.
%Progress in higher-dimensions so far has been mostly theoretical and a  huge gap  remains to be %filled before  having at one's disposal fully satisfactory solutions of practical significance.% One research direction is to invent simplicial complexes of small complexity and easy to compute that still capture the main features of the underlying shape. see Attali and Carlsson. 
 Simplicial complexes can be used to produce fine meshes well suited to scientific computing purposes~\cite{mh-mpc-2002,boissonnat2010meshing}, or much coarser approximations still useful to infer some important features of shapes such as their homology or some local geometric properties~\cite{geometrica-ccl09,nsw-fhm-2008}. 
 A {\em central tenet} in this project is to regard {\em simplicial complexes as a unifying representation of shapes for geometry understanding in higher dimensions}.%\framebox{still open questions}


 Many types of simplicial complexes can be used and a first issue is to choose the appropriate type. This choice depends on the {\em combinatorial and algorithmic complexities} of the complex, as well as on its {\em power to approximate} a shape.
 The choice of a {\em metric} is another  fundamental issue 
that determines the type and quality of an approximation.
 The simple Euclidean distance in the ambient space, while easy to deal with, is often not the right choice.  As already mentioned, when working in high dimensional spaces, the objects of interest have often an intrinsic dimension much smaller than the ambient dimension. It is thus important to exploit the {\em intrinsic geometry} of the objects. Computational intrinsic geometry has not been seriously tackled yet and even a basic question like the existence of  Delaunay triangulations on Riemannian manifolds has been elusive so far.  This question is of utmost importance for anisotropic mesh generation and optimal approximations. Another important situation is when data are not provided as  point clouds in some Euclidean space, but rather as a matrix of pairwise distances (i.e., a {\em discrete metric space}). Although such data may not be sampled from geometric subsets of Riemannian manifolds, they may still carry some interesting topological structures that need to be understood. 
Lastly, let us mention other {\em pseudo-distances} such as Kullback-Leibler, Itakura-Saito or Bregman divergences that may be preferred in information theory, signal and image processing,
% %  In computer  vision, divergences are used throughout all the process of object recognition: to detect features using statistical, algebraic or geometric techniques, and to use these features throughout classifiers.
% %The recognition of objects in potentially complex scenes is a major issue in Computer Vision. 
% % Prominent advances in object recognition integrate two essential stages: the induction of features of limited size, over a potentially huge feature space, and the use of these features for learning and classification. 
 These divergences are usually not genuine distances (they may not be symmetric nor satisfy the triangular inequality) and  geometric data structures and algorithms need to be revisited in this context~\cite{geometrica-6154a}.




\paragraph{Scientific challenge 2 :  Bypassing the curse of dimensionality.} 
Simplicial complexes have been known and studied for a long time in mathematics, but not so much from a computational point of view. This is however a main issue since the complexity of many geometric algorithms and data structures grows exponentially with increasing dimension. It is thus not possible to partition a high dimensional space, which rules out most, if not all, geometric algorithms developed in low dimensions.
Hence, extending computational geometry in high dimensions cannot be done in a straightforward manner and one has to take advantage of additional structural properties of the problem. % This motivated a number of new concepts (e.g., core sets), new algorithmic paradigms (e.g., locality-sensitive hashing) and new analyses (e.g., smoothed analysis).
In this project, we will address the curse of dimensionality by focusing on the inherent structure in the data which we assume to be of relative {\em low intrinsic dimension}.  We will put the emphasis on {\em output-sensitive} algorithms and on {\em average-case} analysis.  First investigations led to very promising results, such as the design of new simplicial complexes with low complexity and approximation algorithms that scale well with the dimension \cite{geometrica-7142i}.


\paragraph{Scientific challenge 3 : Searching for stable models.} 
When dealing with approximations and samples, one needs stability results to ensure that the quantities that are computed are good approximations of the real ones. This is especially true in higher-dimensions where data are usually corrupted by various types of noise.  When the noise magnitude is small, methods have been proposed to robustly estimate topological and geometric properties of shapes~\cite{nsw-tvu-2011}.  The recent and fast developing theory of {\em persistent homology} provides a powerful tool to study  the homology of sampled spaces and to remove topological noise~\cite{eh-ph-2008}.
However, in  many applications, the noise is non local and the previous methods fail.
Recently,  larger families of noise models  have been considered and statistical approaches  have been proposed to provide shape approximations that are stable with respect to   those types of noise~\cite{gpvw-mme-2011}. These methods however do not provide topological guarantees on the approximation and the question of designing computationally tractable estimators converging at an optimal rate remains open. A major challenge is to design  unifying frameworks that {\em embrace statistical approaches and deterministic methods}, and offer topological guarantees.   

The automatic selection of the relevant {\em scales} at which the geometry of data should be considered is another issue.  This can be understood by considering a point set sampling a curve lying on a manifold, say an helix with a small pitch drawn on a cylinder. Depending on the scale, the expected output of a reconstruction algorithm will be an approximation of the curve or of the cylinder. Hence, the analysis process has to be multiscale or at least be able to automatically select the appropriate scale.  Here also, combining persistent homology and other geometric approaches~\cite{geometrica-bgo-09} with statistical techniques would lead to significant progress.

\paragraph{Scientific challenge 4 : Turning theory into practice.}
A major challenge, if not the most important, is to develop {\em theory} that is {\em of practical significance} for applications.   To take up the challenge, we will %  foster a symbiotic relationship between theory and practice, and
undertake the development of a {\em software platform} devoted to geometry understanding in higher dimensions. 
We consider such a platform as central to our research  for three main reasons.  {\em First}, the software platform will allow {\em large scale experimentations}, which is mandatory to design the right models and data structures. We believe that this will revitalize the current theory and open {\em new vistas for research}, both of a practical and a theoretical nature, leading towards a virtuous circle between theory and experimental research. This has proven to be of utmost importance
when developing the CGAL library and will be even more true in high dimensional geometry.

{\em Second}, maintaining such a platform will help further effort and {\em consolidation in the long run}.  Having a library with interoperable modules will allow us to incrementally add more and more sophisticated tools based on solid foundations.  This is consistent with our long-term vision and our conviction that it is only through such a long standing effort that true impact, both theoretical and applied, can be gained.

{\em Third}, the platform will serve as a unique tool to {\em communicate} with the computational geometry community and with researchers from other fields. 
 In return, we will get feedback from practitioners which will help shape the theoretical models and the software platform.

\paragraph{Objectives and research roadmap.}
The ambition of this proposal is to settle the {\em algorithmic
foundations} of geometry understanding in {\em dimensions higher than
3}.  We intend to develop {\em scalable representations in the form of simplicial complexes} and {\em
practical algorithms} to approximate {\em highly nonlinear shapes}, and to
infer geometric and topological properties from data subject to
significant {\em defects} and under {\em realistic conditions}.
As is common in many applications across science and engineering, we
will assume that the objects of interest can be modeled as {\em
  low-dimensional manifolds} embedded in possibly high-dimensional
spaces. By exploiting the {\em intrinsic properties} of the objects,
we will produce intrinsic dimension-sensitive data structures and algorithms
that will {\em break the current computational
bottleneck.}  %models in the form of {\em simplicial complexes}.

% % interest which can be modeled as {\em low-dimensional manifolds} in
% % many applications.  % We % intend to develop {\em practical
% algorithms to approximate % highly nonlinear manifolds, and to infer
% geometric and topological % properties from data subject to
% significant {\em defects} and under % {\em realistic conditions}.  
To reach these objectives, the guiding principle will be to foster a
symbiotic relationship between theory and practice, and to address
{\em fundamental research} issues along three parallel advancing
fronts. We will simultaneously develop {\em mathematical approaches}
providing theoretical guarantees, {\em effective algorithms} that are
amenable to theoretical analysis and rigorous experimental validation,
and {\em perennial software} development. %  We will undertake the
% development of a high-quality open source {\em software platform} to
% implement the most important geometric data structures and algorithms
% at the heart of geometry understanding in higher dimensions. 
 % The central goal of this proposal is to settle the {\em algorithmic foundations of geometry understanding in higher dimensions}.  We aim at processing general highly nonlinear geometries with nontrivial topologies that can be modeled as  {\em low-dimensional manifolds} embedded in possibly high-dimensional spaces. %  Approximation of such manifolds will be in the form of
% % {\em simplicial complexes}.  We intend to bypass the computational bottleneck by exploiting the {\em intrinsic properties} of the objects and to design data structures and algorithms that scale with the intrinsic dimension of the objects of interest rather than with the ambient dimension. We intend to develop {\em practical algorithms} to mesh or reconstruct highly nonlinear manifolds, and to infer geometric and topological properties from data subject to significant {\em defects} and under {\em realistic conditions}. 
% This proposal addresses {\em fundamental
%   research} issues, and its results are expected to serve as a basis
% for groundbreaking advances for {\em applications in scientific computing
% and data analysis}.  % A major outcome of the project will be a
% % high-quality open source software {\em platform} of components
% % implementing the main results. 
% %
% To reach these objectives, the guiding principle  will be to simultaneously
% develop {\em mathematical approaches} providing theoretical
% guarantees, {\em effective algorithms} that are amenable to both
% theoretical analysis and rigorous experimental validation, and {\em
%   perennial software} development.


The proposal is structured into the following four {\em focus areas}  that address the four scientific challenges listed above.
{\bf A1} -- {\em Intrinsic dimension-sensitive data  structures} --  will address Challenges 1 and 2 by extending current knowledge on the combinatorial and algorithmic properties of simplicial complexes. 
  {\bf A2} --  {\em Triangulation of non Euclidean metric spaces} -- will address Challenges 1 and 2 by developing effective algorithms to mesh or reconstruct manifolds equipped with various metrics.   {\bf A3} -- {\em Robust models for geometric and topological inference} -- will address Challenge 3 by providing the crucial  algorithms for topological data analysis.
 {\bf A4} --  {\em  Software platform for geometric understanding in high dimensions} -- will address Challenge~4 by providing the software environment for experimenting with our new data structures and algorithms, for integrating them in a library of interoperable modules, and for diffusing our results to applied fields. 

\paragraph{Risks and feasibility of the project.} 

% The Gudhi proposal may look too ambitious and risky since we want to simultaneously conduct basic research at the best international level and to develop software of the highest quality.
%. Although risky, we argue that this approach is key to success. It can be argued that  my research group Geometrica is the best team worldwide to take up the challenge and to make this project a success. 
 My personal record as well as the record of my research group Geometrica are strong indications of my ability to take up the challenge of simultaneously conducting basic research at the best international level and developing software of the highest quality with good chances of success.  
Geometrica has been at the cutting edge of research in geometric data structures and algorithms~\cite{by-ag-98}, mesh generation~\cite{geometrica-ecg-book}, manifold reconstruction~\cite{geometrica-7142i,geometrica-bgo-09}, geometric inference and computational topology~\cite{geometrica-ccl09,geometrica-cseh-07}. Geometrica is also one of the leader teams of the CGAL project (www.cgal.org).  We were at the source of successful developments in CGAL like interval arithmetics, 2 and 3-dimensional triangulations (now integrated in the heart of Matlab) and meshing packages. We also took a prominent part in the organization of the CGAL project and community, and in the creation of the spinoff GeometryFactory.
%
%The research won't be conducted in isolation.  
We will benefit from our strong collaborations with the best groups in US and  Europe,
most notably with our partners of the ICT Fet-Open project Computational Geometric Learning (CG-Learning) which is closely related to this project (focus, timetable and management are different though).

% Research in computational geometry and topology is very active in  Europe.  The ICT Fet-Open project Computational Geometric Learning (CG-Learning) is closely related to this project. 
% CG-Learning is an exploratory project touching upon several topics and devoting limited ressource to software development. The Gudhi proposal wants to take over the results of CG-Learning
% and to focus on one of the most promissing emerging research direction,  and to go far beyond prototype developments.  



%We will also benefit from our long-standing collaborations in the USA with Stanford university %(Pr. Guibas) and Ohio State university (Pr. Dey) on topics that are related to this project.



\paragraph{New horizons and opportunities.} 



% Computational geometry and topology have produced beautiful pieces of theory and. Although many of these tools are of great potential impact in applications, implementing those ideas in robust and efficient codes will only be possible through a long-standing alliance between mathematical developments, algorithmic design and advanced programming.  This last point is underestimated and implementation is often considered as an engineering activity that can be let to students. I believe that this is a wrong vheartiew that largely explains the current absence of  reliable software toolbox for geometric modeling in higher-dimensions. % My conviction, built after more than ten years of development of CGAL, is  that the suggested workplan is the only way to make  impact in the long run.

Upon successful completion, the Gudhi project will put geometry understanding in higher-dimensions on new theoretical and algorithmic ground. It will help set up a first class research group with a unique spectrum of expertise covering mathematics, algorithm design and software development.  This new project will further strengthen the leadership of Europe in Geometric Computing.  It will also provide an open platform with no equivalent in USA or Asia.  We foresee the project becoming a catalyst for research in high-dimensional geometry inside and outside of the project.  By implementing the most effective techniques in a reliable and scalable way, the platform will open the way to groundbreaking technological advances in scientific computing and data analysis for applications as varied as numerical simulation, neurosciences, astrophysics and molecular biology. We will keep close contacts with a small set of prominent researchers working in those domains and leap on opportunities arising from our new results and tools.  Specific attention will be paid to the analysis of {\em energy landscapes of molecular systems} and to understand the phase space dynamics of {\em cosmic structure formation}.

% {\footnotesize
% \bibliographystyle{abbrv}
% \bibliography{erc}
% }

%\paragraph{References} \mbox{}\\

\begin{thebibliography}{}
\vspace{-2mm}

{\footnotesize

\bibitem[Boissonnat and Ghosh, 2010b]{boissonnat2010meshing}
Boissonnat, J.-D. and Ghosh, A. (2010b).
\newblock Triangulating smooth submanifolds with light scaffolding.
\newblock {\em Mathematics in Computer Science}, 4(4):431--461.
\vspace{-2mm}

\bibitem[Boissonnat et~al., 2009]{geometrica-bgo-09}
Boissonnat, J.-D., Guibas, L.~J., and Oudot, S. (2009).
\newblock Manifold reconstruction in arbitrary dimensions using witness
  complexes.
\newblock {\em Discrete and Computational Geometry}, 42(1):37--70.
\vspace{-2mm}

\bibitem[Chazal et~al., 2011]{ccsm-gipm-2011}
Chazal, F., Cohen-Steiner, D., and M\'erigot, Q. (2011).
\newblock Geometric inference for probability measures.
\newblock {\em Journal on Foundations of Computational Mathematics},
  11(6):733--751.
\vspace{-2mm}

\bibitem[Donoho, 2000]{dld-hdda-2000}
Donoho, D. (2000).
\newblock High-dimensional data analysis: The curses and blessings of
  dimensionality.
\newblock In {\em Mathematical Challenges of the 21st Century}. AMS.
\vspace{-2mm}

\bibitem[Edelsbrunner and Harer, 2010]{hh-ct-2010}
Edelsbrunner, H. and Harer, J. (2010).
\newblock {\em Computational topology}.
\newblock American Mathematical Society.
\vspace{-2mm}

\bibitem[Ghrist, 2008]{rg-bptd-2008}
Ghrist, R. (2008).
\newblock Barcodes: the persistent topology of data.
\newblock {\em Bull. Amer. Math. Soc., 45(1)}, pages 61--75.
\vspace{-2mm}

\bibitem[Niyogi et~al., 2008]{nsw-fhm-2008}
Niyogi, P., Smale, S., and Weinberger, S. (2008).
\newblock Finding the {H}omology of {S}ubmanifolds with {H}igh {C}onfidence
  from {R}andom {S}amples.
\newblock {\em Discrete and Computational Geometry}, 39(1):419--441.

}

\end{thebibliography}
